---
title: "microphase1and2_litter"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

This code works up litter moisture data from the 2022 IRES Microclimate project, both microphase 1 and 2. Original csv file is pulled in, cleaned, analyzed, and graphed.

File written: 13 July 2025, Heather Throop\
Last updated: 2026-01-30 HT - updated file location and directory structure

## Preparation Steps

```{r}
#| echo: false
#| label: load packages 

library(here) 
library(tidyverse)
library(grid)
library(bestNormalize)
library(car)        # for Levene's test
library(ggplot2)
library(emmeans)    # for post hoc Tukey
library(cowplot)
library(ggpattern)
library(png)
library(grid)

sessionInfo()
```

```{r}
# define colors
TYPE_colors <- c("#fa882a","#6e4424","#5dba2b","#166610")
DEPTH_colors <- c("#166610", "tan", "#6e4424")
Rainfall_colors <- c("#d95f02", "#1b9e77", "#7570b3")
TYPE_linetypes <- c("brown grass" = "solid",
              "brown shrub" = "dashed",
              "green grass" = "solid",
              "green shrub" = "dashed")
```

```{r}
# Use the stored config file to define the path for the DERT Data & Metadata folder (in ASU Dropbox) 
DERT_Archives <- config::get()

# Base project paths
data_dir   <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "data")

icon_dir   <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "icons")

stats_dir  <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "output", "stats")

fig_dir    <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "output", "figures")
```

```{r}
# Read in the moon and sun images as grobs
moon_img <- readPNG(file.path(icon_dir, "moon.png"))
moon_grob <- rasterGrob(moon_img, interpolate = TRUE)

sun_img <- readPNG(file.path(icon_dir, "sun.png"))
sun_grob <- rasterGrob(sun_img, interpolate = TRUE)

```

```{r}
#| label: Three-way ANOVA function

run_threeway_anova <- function(df, response_var, factors, 
                               df_name = deparse(substitute(df)), 
                               output_file = NULL) {
  if (length(factors) != 3) {
    stop("You must provide exactly three main effect variable names.")
  }

  # Clean factor names
  factors <- trimws(factors)

  # Ensure all specified variables are factors
  df <- df |>
    mutate(across(all_of(factors), as.factor))

  # Build terms for the model
  all_terms <- c(
    factors,
    paste(factors[1], factors[2], sep = ":"),
    paste(factors[1], factors[3], sep = ":"),
    paste(factors[2], factors[3], sep = ":"),
    paste(factors[1], factors[2], factors[3], sep = ":")
  )

  # Create formula
  formula <- reformulate(termlabels = all_terms, response = response_var)

  # Fit models
  aov_model <- aov(formula, data = df)
  lm_model <- lm(formula, data = df)
  aov_summary <- summary(aov_model)

  # Capture ANOVA output as character
  summary_txt <- capture.output(print(aov_summary))

  # Store post hoc results
  posthoc_results <- list()
  posthoc_txt <- c()

  for (factor in factors) {
    if (nlevels(df[[factor]]) > 1) {
      try({
        emm <- emmeans::emmeans(lm_model, specs = factor)
        pairs_result <- if (nlevels(df[[factor]]) > 2) {
          emmeans::pairs(emm, adjust = "tukey")
        } else {
          "Only 2 levels; no pairwise comparisons needed."
        }

        posthoc_results[[factor]] <- list(
          emm = emm,
          pairs = pairs_result
        )

        # Capture output
        posthoc_txt <- c(
          posthoc_txt,
          paste0("→ ", factor),
          capture.output(print(emm)),
          if (is.character(pairs_result)) pairs_result else capture.output(print(pairs_result)),
          ""
        )
      })
    }
  }

  # Combine text results
  all_txt <- c(
    paste0("\n=== ANOVA Results for Dataframe: ", df_name, " ==="),
    paste0("Response variable: ", response_var),
    paste0("Model formula: ", deparse(formula)),
    "\n--- ANOVA Summary ---",
    summary_txt,
    "\n--- Post Hoc Tests ---",
    posthoc_txt
  )

  # Return results
  return(list(
    aov_model = aov_model,
    lm_model = lm_model,
    anova_summary = aov_summary,
    posthoc = posthoc_results
  ))
}


```

# Microphase 1

## Data Ingestion & Calculations

```{r}
#| label: read in data and calculate derived variables
micro1_df <- read_csv(file.path(data_dir,"microphase1_litter_data.csv"))
micro1_df <- na.omit(micro1_df) # remove lines with NAs

## data preparations
# change rainfall to categorical (these are treatment names)
micro1_df$Rainfall <- as.factor(micro1_df$Rainfall) 
# use 'as_datetime' from lubridate to change Datetime to POSIXct format
micro1_df$Datetime <- as_datetime(micro1_df$Datetime, format = "%m/%d/%y %H:%M", tz = "Africa/Johannesburg")

# Define the experiment start time in Johannesburg time zone
start_time <- as.POSIXct("2022-07-24 11:00:00", tz = "Africa/Johannesburg")

# Compute elapsed time in hours
micro1_df <- micro1_df |>
  mutate(elapsed_time = as.numeric(difftime(Datetime, start_time, units = "hours")))

# define night (6 pm to 6 am) periods during the experiment (relative to elapsed time)
night1_start_m1 <- 7
night1_end_m1 <- 19
night2_start_m1 <- 31
night2_end_m1 <- 43

# recode species names to functional types, calling recode explictly as it seems to be masked
micro1_df$Species <- dplyr::recode(micro1_df$Species,
                                   "SAPE" = "shrub",
                                   "STSA" = "grass")
# recode TYPE to condition and functional types, calling recode explictly as it seems to be masked
micro1_df$TYPE <- dplyr::recode(micro1_df$TYPE,
                                   "BSP" = "brown shrub",
                                   "BSS" = "brown grass",
                                   "GSP" = "green shrub",
                                   "GSS" = "green grass")

# force the order for TYPE so color assignments are in the same order
micro1_df <- micro1_df |> 
  mutate(TYPE = factor(TYPE, 
                       levels = c("brown grass", "brown shrub", 
                                  "green grass", "green shrub")))


##calculate derived variables
# calculate litter wet mass, dry mass, and percent moisture at collection
micro1_df <- micro1_df |> 
  mutate(LitterWetMass = Sample_mass - SandMasswBag, # sample wet mass at collection
         LitterDryMass = Drymasswbag - D_bagMasswSand, # sample dry mass at collection
         LitterMoisturePct = ((LitterWetMass - LitterDryMass) / LitterDryMass) *100)   

# explore LitterMoisturePct data with a histogram
hist(micro1_df$LitterMoisturePct,
     main = "Histogram of Litter Moisture (%)",
     xlab = "WHC (% by mass)",
     col = "skyblue",
     border = "white",
     breaks = 20)

# There are two outlier where LitterMoisturePct is over 100 while other values in the same group are ~5 (for elapsed_time = 12) and 85 (for elapsed_time = 48). Remove these.
micro1_df <- micro1_df |>
  filter(!(Species == "grass" &
           Littercond == "Green" &
           elapsed_time == 12 &
           LitterMoisturePct > 100),
         !(Species == "shrub" &
           Littercond == "Brown" &
           elapsed_time == 48 &
           LitterMoisturePct > 100))
```

## Litter Data Plotting

```{r}
#| label: calculate means

# calculate mean values for line graphs 
micro1_means <- micro1_df |> 
  group_by(Species, Littercond, Rainfall, TYPE, elapsed_time) |> 
  summarise(LitterMoisture_mean=mean(LitterMoisturePct, na.rm=TRUE), 
            n=n(), 
            sd=sd(LitterMoisturePct, na.rm=TRUE), 
            se=sd/sqrt(n))
```

```{r}
#| label: litter moisture plot by species and condition - NOT USED

micro1_litter_p <- micro1_means |>
  ggplot(aes(x = elapsed_time, y = LitterMoisture_mean, 
                       color = as.factor(Rainfall),
                       shape = as.factor(Rainfall),
                       linetype = as.factor(Rainfall),
                       group = Rainfall)) +
  annotate("rect", xmin = 35, xmax = 45, ymin = -5, ymax = Inf, 
           alpha = 0.1, fill = "skyblue2") +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = LitterMoisture_mean - se, 
                    ymax = LitterMoisture_mean + se), width = 0.5) +
  facet_grid(rows = vars(Littercond), cols = vars(Species)) +
  # Custom colors, shapes, and line types for Rainfall
  scale_color_manual(values = c("0" = "red", "5" = "green", "10" = "blue")) +
  scale_shape_manual(values = c("0" = 16,  # filled circle
                                "5" = 15,  # filled square
                                "10" = 17  # filled triangle
                                )) +
  scale_linetype_manual(values = c("0" = "solid",
                                   "5" = "dotted",
                                   "10" = "dashed")) +
  labs(x = "Elapsed Time",
       y = "Litter Moisture (% by mass)",
       color = "",
       shape = "",
       linetype = "") +
  theme_classic(base_size = 16) +
 theme(
  panel.spacing = unit(1, "lines"),
  strip.background = element_rect(fill = "gray80", color = "black"),
  strip.text = element_text(color = "black"),
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)) +
  scale_y_continuous(limits = c(-5, NA)) +
  theme(
  legend.position = c(0.05, 0.2),  # x and y between 0 (left/bottom) and 1 (right/top)
  legend.justification = c("left", "bottom"),
  legend.background = element_rect(fill = alpha('white', 0.6), color = NA))
micro1_litter_p


# save the plot
ggsave(filename = 
       file.path(fig_dir, "microphase1_littermoisture.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

```{r}
#| label: litter moisture by rainfall 

micro1_litter_p2 <- micro1_means |>
    mutate(
    Rainfall = fct_recode(Rainfall,
      "A. No Addition"      = "0",
      "B. 5 mm Addition"   = "5",
      "C. 10 mm Addition"   = "10"
    )
  ) |>
  ggplot(aes(x = elapsed_time, y = LitterMoisture_mean,
             color = TYPE,
             shape = TYPE,
             linetype = TYPE,
             group = TYPE)) +
  # add blue box for fog
  annotate("rect", xmin = 33, xmax = 44, 
           ymin = -5, ymax = Inf, alpha = 0.5, fill = "#54b8e3") +
  # add fog label
  # annotate("text", x = 38.5, y = 130, label = "Fog", color = "#54b8e3", size = 5) +
  # Add shaded night areas
  annotate("rect",
         xmin = night1_start_m1, xmax = night1_end_m1, 
         ymin = -5, ymax = Inf,
         alpha = 0.4, fill = "darkgrey",
         colour = "black", linewidth = 0.3) +
  annotate("rect", 
         xmin = night2_start_m1, xmax = night2_end_m1, 
         ymin = -5, ymax = Inf, 
         alpha = 0.4, fill = "darkgrey",
         colour = "black", linewidth = 0.3) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = LitterMoisture_mean - se, 
                    ymax = LitterMoisture_mean + se), 
                width = 0.5,
                linetype = "solid") +
  facet_grid(. ~ Rainfall) +
  scale_color_manual(values = TYPE_colors) +
  scale_shape_manual(values = c(16, 17, 16, 17)) +
  scale_linetype_manual(values = TYPE_linetypes) + # custom linetypes
  labs(x = NULL,
       y = "Litter Moisture (%)",
       color = NULL) +   # remove the title above legend
  guides(
    color = guide_legend(override.aes = list(shape = c(16, 17, 16, 17), linetype = c("solid", "dashed", "solid", "dashed"))),
    shape = "none",      # <-- hide shape legend
    linetype = "none"    # <-- hide linetype legend
  ) +
  theme_classic(base_size = 16) +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.background = element_rect(fill = "gray80", color = "black"),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position = "top",
    legend.key.width = unit(1.5, "cm"),
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.key.size = unit(0.8, "lines"),
    legend.spacing.x = unit(0.3, "cm"),
    legend.text = element_text(size = 12)
  ) +
  scale_y_continuous(limits = c(-5, 155), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 48)) +
  # Add moon images in night areas
  annotation_custom(grob = moon_grob, 
        xmin = 8, xmax = 18, ymin = 137, ymax = 147) +
  # Add sun images in day areas
  annotation_custom(grob = sun_grob, 
        xmin = 19, xmax = 31, ymin = 136, ymax = 148) 

micro1_litter_p2

# save the plot
ggsave(filename = 
       file.path(fig_dir, "microphase1_littermoisture_by_ppt.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

## iButton Data Plotting

```{r}
#| label: read in L1 iButton CSV
Microphase1_iButtons_df <- read_csv(file.path(data_dir,"microphase1_iButtons_L1.csv"))

# change to a factor and change capitalization for consistency
Microphase1_iButtons_df$Rainfall <- as.factor(Microphase1_iButtons_df$rainfall)

# add new column for TYPE that combines species and condition (to be comparable with litter data)
Microphase1_iButtons_df <- Microphase1_iButtons_df |>
  mutate(TYPE = case_when(
    species == "SP" & condition == "B" ~ "brown shrub",
    species == "SS" & condition == "B" ~ "brown grass",
    species == "SP" & condition == "G" ~ "green shrub",
    species == "SS" & condition == "G" ~ "green grass",
    TRUE ~ NA_character_
  ))

# force the order for TYPE so color assignments are in the same order
Microphase1_iButtons_df <- Microphase1_iButtons_df |> 
  mutate(TYPE = factor(TYPE, 
                       levels = c("brown grass", "brown shrub", 
                                  "green grass", "green shrub")))

```

```{r}
#| label: bin iButton temperature into hourly means

Microphase1_iButtons_hourly_df <- Microphase1_iButtons_df |>
  # create a "bin" around each whole number of elapsed_time
  mutate(elapsed_bin = round(elapsed_time)) |>
  group_by(quadrat, Rainfall, species, condition,TYPE, elapsed_bin) |>
  summarise(
    temp_hourly_mean = mean(temperature, na.rm = TRUE),
    N = n()
  ) |>
  group_by(Rainfall, species, condition, TYPE, elapsed_bin) |>
  summarise(
    temperature = mean(temp_hourly_mean, na.rm = TRUE),
    N = n(),
    sd_temp = sd(temp_hourly_mean, na.rm = TRUE)
  )

```

```{r}
#| label: iButton temperature by rainfall 

# NOT USING - USE BINNED DATA (Below)

micro1_temp_p1 <- Microphase1_iButtons_df |>
    mutate(
    Rainfall = fct_recode(Rainfall,
      "D. No Addition"      = "0",
      "E. 5 mm Addition"   = "5",
      "F. 10 mm Addition"   = "10"
    )
  ) |>
  ggplot(aes(x = elapsed_time, y = temperature,
             color = TYPE,
             shape = TYPE,
             linetype = TYPE,
             group = TYPE)) +
  # add blue box for fog
  annotate("rect", xmin = 33, xmax = 44, 
           ymin = -Inf, ymax = Inf, alpha = 0.5, fill = "#54b8e3") +
  # add fog label
  # annotate("text", x = 38.5, y = 130, label = "Fog", color = "#54b8e3", size = 5) +
  # Add shaded night areas
  annotate("rect",
         xmin = night1_start_m1, xmax = night1_end_m1, 
         ymin = -Inf, ymax = Inf,
         alpha = 0.4, fill = "darkgrey",
         colour = "black", linewidth = 0.3) +
  annotate("rect", 
         xmin = night2_start_m1, xmax = night2_end_m1, 
         ymin = -Inf, ymax = Inf, 
         alpha = 0.4, fill = "darkgrey",
         colour = "black", linewidth = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.2, linewidth = 0.5) +
  facet_grid(. ~ Rainfall) +
  scale_color_manual(values = TYPE_colors) +
  scale_linetype_manual(values = TYPE_linetypes) + # custom linetypes
  scale_x_continuous(limits = c(0, 48)) +
  scale_y_continuous(limits = c(5, 45), expand = c(0, 0)) +
  labs(x = "Elapsed Time (hours)",
       y = "Litter Temp. (°C)",
       color = "",
       shape = "",
       linetype = "") +
  theme_classic(base_size = 16) +
  theme(
    panel.spacing = unit(1, "lines"),
    legend.position = "top",
    strip.background = element_rect(fill = "gray80", color = "black"),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
micro1_temp_p1

# save the plot
ggsave(filename = 
       file.path(fig_dir, "microphase1_temperature_by_ppt.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

```{r}
#| label: iButton temperature by rainfall - binned hourly data

micro1_temp_binned_p1 <- Microphase1_iButtons_hourly_df |>
    mutate(
    Rainfall = fct_recode(Rainfall,
      "D. No Addition"      = "0",
      "E. 5 mm Addition"   = "5",
      "F. 10 mm Addition"   = "10"
    )
  ) |>
  ggplot(aes(x = elapsed_bin, y = temperature,
             color = TYPE,
             shape = TYPE,
             linetype = TYPE,
             group = TYPE)) +
  # add blue box for fog
  annotate("rect", xmin = 33, xmax = 44, 
           ymin = -Inf, ymax = Inf, alpha = 0.5, fill = "#54b8e3") +
  # add fog label
  # annotate("text", x = 38.5, y = 130, label = "Fog", color = "#54b8e3", size = 5) +
  # Add shaded night areas
  annotate("rect",
         xmin = night1_start_m1, xmax = night1_end_m1, 
         ymin = -Inf, ymax = Inf,
         alpha = 0.4, fill = "darkgrey",
         colour = "black", linewidth = 0.3) +
  annotate("rect", 
         xmin = night2_start_m1, xmax = night2_end_m1, 
         ymin = -Inf, ymax = Inf, 
         alpha = 0.4, fill = "darkgrey",
         colour = "black", linewidth = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.2, linewidth = 0.5) +
  facet_grid(. ~ Rainfall) +
  scale_color_manual(values = TYPE_colors) +
  scale_linetype_manual(values = TYPE_linetypes) + # custom linetypes
  scale_x_continuous(limits = c(0, 48)) +
  scale_y_continuous(limits = c(5, 45), expand = c(0, 0)) +
  labs(x = "Elapsed Time (hours)",
       y = "Litter Temp. (°C)",
       color = "",
       shape = "",
       linetype = "") +
  theme_classic(base_size = 16) +
  theme(
    panel.spacing = unit(1, "lines"),
    legend.position = "top",
    strip.background = element_rect(fill = "gray80", color = "black"),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
micro1_temp_binned_p1

# save the plot
ggsave(filename = 
       file.path(fig_dir, "microphase1_temperature_by_ppt_binned.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

## Combine litter moisture and temperature plots

```{r}
#| label: Combine plots (FIG3)

# Extract the legend grob from the first plot

# supress legends in plot 2, use plot 1 legend in combined plot
plot1_legend <- micro1_litter_p2 + theme(legend.position = "top")
plot2_no_legend <- micro1_temp_binned_p1 + theme(legend.position = "none")

combined_plots <- plot_grid(plot1_legend, plot2_no_legend, ncol = 1, align = "v", rel_heights = c(1, 1))

final_plot <- ggdraw() +
  draw_plot(combined_plots, x = 0, y = 0, width = 1, height = 1)
final_plot

# save the plot
ggsave(filename = 
       file.path(fig_dir, "Fig3.microphase1_litter_and_temp.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)

```

## Litter Moisture ANOVAs

```{r}
#| label: check ANOVA assumptions

# Filter and prepare data
micro1_df <- micro1_df |>
  mutate(Group = interaction(Species, Littercond, Rainfall))

# Step 1: Check initial normality
hist(micro1_df$LitterMoisturePct, main = "Histogram Litter Moisture", xlab = "response var")
qqnorm(micro1_df$LitterMoisturePct); qqline(micro1_df$LitterMoisturePct)

shapiro_raw <- shapiro.test(micro1_df$LitterMoisturePct)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Step 2: Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(micro1_df$LitterMoisturePct, allow_orderNorm = FALSE)
  micro1_df <- micro1_df |>
    mutate(LitterMoisturePct_trans = predict(bn_obj, newdata = LitterMoisturePct))
  used_var <- "LitterMoisturePct_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(micro1_df$LitterMoisturePct_trans, 
       main = "Histogram of Transformed Moisture", xlab = "response var_trans")
  qqnorm(micro1_df$LitterMoisturePct_trans); qqline(micro1_df$LitterMoisturePct_trans)
  
  shapiro_trans <- shapiro.test(micro1_df$LitterMoisturePct_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "LitterMoisturePct"
  transform_used <- "None"
}

# Step 3: Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = micro1_df)

# Step 4: Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)
```

```{r}
#| label: three-way ANOVAs at key time points

# uses the ANOVA function laoded in the preparations steps chunk

# Separate into dfs for the key time points (8, 20, 32, 44)
micro1_8h_df <- micro1_df |>
  filter(elapsed_time == "8")

micro1_20h_df <- micro1_df |>
  filter(elapsed_time == "20")

micro1_32h_df <- micro1_df |>
  filter(elapsed_time == "32")

micro1_44h_df <- micro1_df |>
  filter(elapsed_time == "44")

# Three-way ANOVA models using the function defined in the preparation steps
result_8h <- run_threeway_anova(
  df = micro1_8h_df,
  response_var = used_var,
  factors = c("Species", "Littercond", "Rainfall")
)
print(result_8h)
# Save output to a text file
sink(file.path(stats_dir, "anova_summary_micro1_8h.txt"))
cat("\n=== ANOVA Summary ===\n")
print(result_8h)
sink()

result_20h <- run_threeway_anova(
  df = micro1_20h_df,
  response_var = used_var,
  factors = c("Species", "Littercond", "Rainfall")
)
print(result_20h)
# Save output to a text file
sink(file.path(stats_dir, "anova_summary_micro1_20h.txt"))
cat("\n=== ANOVA Summary ===\n")
print(result_20h)
sink()

result_32h <- run_threeway_anova(
  df = micro1_32h_df,
  response_var = used_var,
  factors = c("Species", "Littercond", "Rainfall")
)
print(result_32h)
# Save output to a text file
sink(file.path(stats_dir, "anova_summary_micro1_32h.txt"))
cat("\n=== ANOVA Summary ===\n")
print(result_32h)
sink()

result_44h <- run_threeway_anova(
  df = micro1_44h_df,
  response_var = used_var,
  factors = c("Species", "Littercond", "Rainfall")
)
print(result_44h)
# Save output to a text file
sink(file.path(stats_dir, "anova_summary_micro1_44h.txt"))
cat("\n=== ANOVA Summary ===\n")
print(result_44h)
sink()
```

# Microphase 2

## Data Ingestion & Calculations

```{r}
#| label: read in data and calculate derived variables
micro2_df <- read_csv(file.path(data_dir,"microphase2_litter_data.csv"))
micro2_df <- na.omit(micro2_df) # remove lines with NAs

## data preparations
# change rainfall and DEPTH to categorical (these are treatment names)
micro2_df$Rainfall <- as.factor(micro2_df$Rainfall) 
micro2_df$DEPTH <- as.factor(micro2_df$DEPTH)
# change Wmass, Dmass, and Moist to numeric (these are chr)
micro2_df$LitterWetMass <- as.numeric(micro2_df$Wmass) 
micro2_df$LitterDryMass <- as.numeric(micro2_df$Dmass) 
micro2_df$LitterMoisturePct <- as.numeric(micro2_df$Moist) 
# use 'as_datetime' from lubridate to change Datetime to POSIXct format
micro2_df$Datetime <- as_datetime(micro2_df$Datetime, format = "%m/%d/%y %H:%M", tz = "Africa/Johannesburg")

# Define the experiment start time in Johannesburg time zone
start_time_m2 <- as.POSIXct("2022-07-30 12:00:00", tz = "Africa/Johannesburg")

# Compute elapsed time in hours
micro2_df <- micro2_df |>
  mutate(elapsed_time = as.numeric(difftime(Datetime, start_time_m2, units = "hours")))

# Define night start and end (6 pm to 6 am) based on elapsed time
night1_start_m2 <- 6
night1_end_m2 <- 18
night2_start_m2 <- 30
night2_end_m2 <- 42

# explore LitterMoisturePct data with a histogram
hist(micro2_df$LitterMoisturePct,
     main = "Histogram of Litter Moisture (%)",
     xlab = "Litter Moisture (% by mass)",
     col = "skyblue",
     border = "white",
     breaks = 20)

# There are five outliers where LitterMoisturePct is less than -10%. Remove these.
micro2_df <- micro2_df |>
  filter(!(LitterMoisturePct < -10))
# filter out outliers in 1 mm DEPTH, 0 mm Rainfall with LitterMoisturePct >110%
micro2_df <- micro2_df |>
  filter(!(DEPTH == "1" & Rainfall == "0" & LitterMoisturePct > 110))
# filter out outliers for rainfall 0 mm, elapsed time 42 h 
micro2_df <- micro2_df |>
  filter(!(Rainfall == "0" & elapsed_time == 42 & LitterMoisturePct > 35))

```

## Litter Data Plotting

```{r}
#| label: calculate means

# calculate mean values for line graphs 
micro2_means <- micro2_df |> 
  group_by(DEPTH, Rainfall, elapsed_time) |> 
  summarise(LitterMoisture_mean=mean(LitterMoisturePct, na.rm=TRUE), 
            n=n(), 
            sd=sd(LitterMoisturePct, na.rm=TRUE), 
            se=sd/sqrt(n))
```

```{r}
#| label: plot litter moisture by rainfall and depth

micro2_litter_p1 <- micro2_means |>
  mutate(
    DEPTH = fct_recode(DEPTH,
      "A. Surface"      = "0",
      "B. 1 cm Depth"   = "1",
      "C. 3 cm Depth"   = "3"
    ),
    Rainfall = fct_recode(Rainfall,
      "0 mm" = "0",
      "5 mm" = "5",
      "10 mm" = "10")
  ) |>
  ggplot(aes(x = elapsed_time, y = LitterMoisture_mean,
             color = Rainfall,
             shape = Rainfall,
             linetype = Rainfall,
             group = Rainfall)) +
 # add blue box for fog
  annotate("rect", xmin = 18, xmax = 19, 
           ymin = -Inf, ymax = Inf, alpha = 0.5, fill = "#54b8e3") +
   # add fog label
 # annotate("text", x = 19, y = 65, label = "Fog", color = "#54b8e3", size = 5) +
  # Add shaded night areas
  annotate("rect", xmin = night1_start_m2, xmax = night1_end_m2, 
           ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "grey",
           colour = "black", linewidth = 0.3) +
  annotate("rect", xmin = night2_start_m2, xmax = night2_end_m2, 
           ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "grey",
           colour = "black", linewidth = 0.3) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = LitterMoisture_mean - se, 
        ymax = LitterMoisture_mean + se),
    width = 0.5,
    linetype = "solid",
    show.legend = FALSE) +
  facet_grid(. ~ DEPTH) +
  labs(x = NULL,
       y = "Litter Moisture (%)",
       color = NULL) +   # no legend title
scale_color_manual(values = Rainfall_colors) +
scale_shape_manual(values = c(16, 17, 18)) +
scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
guides(
  color = guide_legend(
    override.aes = list(
      shape = c(16, 17, 18),
      linetype = c("solid", "dashed", "dotted")
    )
  ),
  shape = "none",
  linetype = "none"
)+
theme(legend.title = element_blank()) +

  theme_classic(base_size = 16) +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.background = element_rect(fill = "gray80", color = "black"),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position = "top",
  #  legend.background = element_rect(fill = alpha("white", 0.7), color = "black"),
    legend.key.width = unit(1.5, "cm"),
    legend.justification = c("center", "top"),
    legend.background = element_blank(),
    legend.key.size = unit(0.8, "lines"),
    legend.spacing.x = unit(0.3, "cm"),
    legend.text = element_text(size = 12)
  ) +
  scale_y_continuous(limits = c(0, 80), expand = c(0, 0)) +
  scale_x_continuous(limits = c(0, 44)) +
  
  # Add moon images in night areas
  annotation_custom(grob = moon_grob, xmin = 10, xmax = 15, ymin = 70, ymax = 75) +
  annotation_custom(grob = moon_grob, xmin = 34, xmax = 39, ymin = 70, ymax = 75) +
  
  # Add sun images in day areas
  annotation_custom(grob = sun_grob, xmin = 21, xmax = 26, ymin = 70, ymax = 75) 

micro2_litter_p1

# save the plot
ggsave(filename = 
       file.path(fig_dir, "microphase2_littermoisture_by_ppt_and_depth.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

## iButton Data Plotting

```{r}
#| label: read in L1 iButton CSV

### NOTE: time in the datetime column is two hours earlier than the time column. I'm assuming for now that datetime is correct, but need to check this carefully.

Microphase2_iButtons_df <- read_csv(file.path(data_dir,"microphase2_iButtons_L1.csv"))

# remove duplicate rows
Microphase2_iButtons_df <- Microphase2_iButtons_df |>
  distinct()

# change to a factor and change capitalization for consistency
Microphase2_iButtons_df$Rainfall <- as.factor(Microphase2_iButtons_df$rainfall)

# add new column for DEPTH that uses the same classification as the litter data
Microphase2_iButtons_df <- Microphase2_iButtons_df |>
  mutate(DEPTH = case_when(
    depth == "D0" ~ "0",
    depth == "D1"  ~ "1",
    depth == "D3" ~ "3",
    TRUE ~ NA_character_
  ))
```

```{r}
#| label: bin iButton temperature into hourly means

Microphase2_iButtons_hourly_df <- Microphase2_iButtons_df |>
  # create a "bin" around each whole number of elapsed_time
  mutate(elapsed_bin = round(elapsed_time)) |>
  group_by(DEPTH, quadrat, Rainfall, elapsed_bin) |>
  summarise(
    temp_hourly_mean = mean(temperature, na.rm = TRUE),
    N = n()
  ) |>
  group_by(DEPTH, Rainfall, elapsed_bin) |>
  summarise(
    temperature = mean(temp_hourly_mean, na.rm = TRUE),
    N = n(),
    sd_temp = sd(temp_hourly_mean, na.rm = TRUE)
  )


```

```{r}
#| label: plot iButton temperature by rainfall - loess smooth
# currently using the binned data (below)

micro2_temp_p1 <- Microphase2_iButtons_df |>
    mutate(
    Rainfall = fct_recode(Rainfall,
      "0 mm" = "0",
      "5 mm" = "5",
      "10 mm" = "10"),
    DEPTH = fct_recode(DEPTH,
      "D. Surface"      = "0",
      "E. 1 cm Depth"   = "1",
      "F. 3 cm Depth"   = "3"
    )
  ) |>
  ggplot(aes(x = elapsed_time, y = temperature,
             color = Rainfall,
             shape = Rainfall,
             linetype = Rainfall,
             group = Rainfall)) +
 # add blue box for fog
  annotate("rect", xmin = 18, xmax = 19, 
           ymin = -Inf, ymax = Inf, alpha = 0.5, fill = "#54b8e3") +
  # Add shaded night areas
  annotate("rect", xmin = night1_start_m2, xmax = night1_end_m2, 
           ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "grey",
           colour = "black", linewidth = 0.3) +
  annotate("rect", xmin = night2_start_m2, xmax = night2_end_m2, 
           ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "grey",
           colour = "black", linewidth = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.2, linewidth = 0.5) +
  facet_grid(. ~ DEPTH) +
  scale_color_manual(values = Rainfall_colors) +
#  scale_shape_manual(values = c(16, 17, 18)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  scale_x_continuous(limits = c(0, 44)) +
  scale_y_continuous(limits = c(0, 35), expand = c(0, 0)) +
  labs(x = "Elapsed Time (hours)",
       y = "Litter Temp. (°C)",
       color = "",
       shape = "",
       linetype = "") +
  theme_classic(base_size = 16) +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.background = element_rect(fill = "gray80", color = "black"),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position = "top",
    legend.justification = c("center","top")
  )
micro2_temp_p1

# save the plot
#ggsave(filename = 
#       here("output","figures", "microphase2_temperature_by_ppt.jpg"),
#       device = "jpeg",
#       width = 8, height = 6, dpi = 300)
```

```{r}
#| label: plot iButton temperature by rainfall - loess smooth hourly binned
# this is the same as above but uses hourly binned data to get a smoother line - use as of 2025-08-22

micro2_temp_p1c <- Microphase2_iButtons_hourly_df |>
    mutate(
    Rainfall = fct_recode(Rainfall,
      "0 mm" = "0",
      "5 mm" = "5",
      "10 mm" = "10"),
    DEPTH = fct_recode(DEPTH,
      "D. Surface"      = "0",
      "E. 1 cm Depth"   = "1",
      "F. 3 cm Depth"   = "3"
    )
  ) |>
  ggplot(aes(x = elapsed_bin, y = temperature,
             color = Rainfall,
             shape = Rainfall,
             linetype = Rainfall,
             group = Rainfall)) +
 # add blue box for fog
  annotate("rect", xmin = 18, xmax = 19, 
           ymin = -Inf, ymax = Inf, alpha = 0.5, fill = "#54b8e3") +
  # Add shaded night areas
  annotate("rect", xmin = night1_start_m2, xmax = night1_end_m2, 
           ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "grey",
           colour = "black", linewidth = 0.3) +
  annotate("rect", xmin = night2_start_m2, xmax = night2_end_m2, 
           ymin = -Inf, ymax = Inf, 
           alpha = 0.2, fill = "grey",
           colour = "black", linewidth = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.2, linewidth = 0.5) +
  facet_grid(. ~ DEPTH) +
  scale_color_manual(values = Rainfall_colors) +
#  scale_shape_manual(values = c(16, 17, 18)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  scale_x_continuous(limits = c(0, 44)) +
  scale_y_continuous(limits = c(0, 35), expand = c(0, 0)) +
  labs(x = "Elapsed Time (hours)",
       y = "Litter Temp. (°C)",
       color = "",
       shape = "",
       linetype = "") +
  theme_classic(base_size = 16) +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.background = element_rect(fill = "gray80", color = "black"),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    legend.position = "top",
    legend.justification = c("center","top")
  )
micro2_temp_p1c

# save the plot
ggsave(filename = 
      file.path(fig_dir, "microphase2_temperature_by_ppt.jpg"),
      device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

## Combine litter moisture and temperature plots

```{r}
#| label: Combine plots (FIG 5)

# supress legends in plot 2, use plot 1 legends for combined plot
plot1_legend <- micro2_litter_p1 + theme(legend.position = "top")
plot2_no_legend <- micro2_temp_p1c + theme(legend.position = "none")

combined_plots <- plot_grid(plot1_legend, plot2_no_legend, ncol = 1, align = "v", rel_heights = c(1, 1))

final_plot <- ggdraw() +
  draw_plot(combined_plots, x = 0, y = 0, width = 1, height = 1)
final_plot

# save the plot
ggsave(filename = 
       file.path(fig_dir, "Fig5_microphase2_litter_and_temp.jpg"),
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

# Litter Moisture ANOVAs

```{r}
#| label: check ANOVA assumptions

# Filter and prepare data
micro2_df <- micro2_df |>
  mutate(Group = interaction(DEPTH, Rainfall))

# Step 1: Check initial normality
hist(micro2_df$LitterMoisturePct, main = "Histogram Litter Moisture", xlab = "response var")
qqnorm(micro2_df$LitterMoisturePct); qqline(micro2_df$LitterMoisturePct)

shapiro_raw <- shapiro.test(micro2_df$LitterMoisturePct)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Step 2: Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(micro2_df$LitterMoisturePct, allow_orderNorm = FALSE)
  micro2_df <- micro2_df |>
    mutate(LitterMoisturePct_trans = predict(bn_obj, newdata = LitterMoisturePct))
  used_var <- "LitterMoisturePct_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(micro2_df$LitterMoisturePct_trans, 
       main = "Histogram of Transformed Moisture", xlab = "response var_trans")
  qqnorm(micro2_df$LitterMoisturePct_trans); qqline(micro2_df$LitterMoisturePct_trans)
  
  shapiro_trans <- shapiro.test(micro2_df$LitterMoisturePct_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "LitterMoisturePct"
  transform_used <- "None"
}

# Step 3: Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = micro2_df)

# Step 4: Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)
```

```{r}
#| label: two-way ANOVAs at key time points

# uses the ANOVA function loaded in the preparations steps chunk
# Separate into dfs for the key time points (6, 18, 30, 42)
micro2_6h_df <- micro2_df |>
  filter(elapsed_time == "6")
micro2_18h_df <- micro2_df |>
  filter(elapsed_time == "18")
micro2_30h_df <- micro2_df |>
  filter(elapsed_time == "30")
micro2_42h_df <- micro2_df |>
  filter(elapsed_time == "42")

# Two-way ANOVA models 

# 6h
anova_model <- aov(reformulate(c("DEPTH", "Rainfall",
                                  "DEPTH:Rainfall"),
                                used_var),
                   data = micro2_6h_df)
anova_summary <- summary(anova_model)

# Create an lm model for emmeans compatibility
lm_model <- lm(
  reformulate(c("DEPTH", "Rainfall", "DEPTH:Rainfall"), used_var),
  data = micro2_6h_df
)
anova_summary <- summary(anova_model)

# Compute emmeans and pairwise comparisons
emm_depth <- emmeans(lm_model, ~ DEPTH)
emm_rain <- emmeans(lm_model, ~ Rainfall)

# Pairwise comparisons with Tukey adjustment
pairs_depth <- pairs(emm_depth, adjust = "tukey")
pairs_rain <- pairs(emm_rain, adjust = "tukey")

# save everything as a text file
sink(file.path(stats_dir, "anova_summary_micro2_6h.txt"))

cat("=== ANOVA Summary ===\n")
print(anova_summary)

cat("\n=== Post Hoc: DEPTH ===\n")
print(emm_depth)
cat("\n--- Pairwise Comparisons (DEPTH, Tukey) ---\n")
print(pairs_depth)

cat("\n=== Post Hoc: Rainfall ===\n")
print(emm_rain)
cat("\n--- Pairwise Comparisons (Rainfall, Tukey) ---\n")
print(pairs_rain)

sink()

# 18h
anova_model <- aov(reformulate(c("DEPTH", "Rainfall",
                                  "DEPTH:Rainfall"),
                                used_var),
                   data = micro2_18h_df)
anova_summary <- summary(anova_model)

# Create an lm model for emmeans compatibility
lm_model <- lm(
  reformulate(c("DEPTH", "Rainfall", "DEPTH:Rainfall"), used_var),
  data = micro2_18h_df
)
anova_summary <- summary(anova_model)

# Compute emmeans and pairwise comparisons
emm_depth <- emmeans(lm_model, ~ DEPTH)
emm_rain <- emmeans(lm_model, ~ Rainfall)

# Pairwise comparisons with Tukey adjustment
pairs_depth <- pairs(emm_depth, adjust = "tukey")
pairs_rain <- pairs(emm_rain, adjust = "tukey")

# save everything as a text file
sink(file.path(stats_dir, "anova_summary_micro2_18h.txt"))

cat("=== ANOVA Summary ===\n")
print(anova_summary)

cat("\n=== Post Hoc: DEPTH ===\n")
print(emm_depth)
cat("\n--- Pairwise Comparisons (DEPTH, Tukey) ---\n")
print(pairs_depth)

cat("\n=== Post Hoc: Rainfall ===\n")
print(emm_rain)
cat("\n--- Pairwise Comparisons (Rainfall, Tukey) ---\n")
print(pairs_rain)

sink()

#30 h
anova_model <- aov(reformulate(c("DEPTH", "Rainfall",
                                  "DEPTH:Rainfall"),
                                used_var),
                   data = micro2_30h_df)
anova_summary <- summary(anova_model)

# Create an lm model for emmeans compatibility
lm_model <- lm(
  reformulate(c("DEPTH", "Rainfall", "DEPTH:Rainfall"), used_var),
  data = micro2_30h_df
)
anova_summary <- summary(anova_model)

# Compute emmeans and pairwise comparisons
emm_depth <- emmeans(lm_model, ~ DEPTH)
emm_rain <- emmeans(lm_model, ~ Rainfall)

# Pairwise comparisons with Tukey adjustment
pairs_depth <- pairs(emm_depth, adjust = "tukey")
pairs_rain <- pairs(emm_rain, adjust = "tukey")

# save everything as a text file
sink(file.path(stats_dir, "anova_summary_micro2_30h.txt"))

cat("=== ANOVA Summary ===\n")
print(anova_summary)

cat("\n=== Post Hoc: DEPTH ===\n")
print(emm_depth)
cat("\n--- Pairwise Comparisons (DEPTH, Tukey) ---\n")
print(pairs_depth)

cat("\n=== Post Hoc: Rainfall ===\n")
print(emm_rain)
cat("\n--- Pairwise Comparisons (Rainfall, Tukey) ---\n")
print(pairs_rain)

sink()

#42 h
anova_model <- aov(reformulate(c("DEPTH", "Rainfall",
                                  "DEPTH:Rainfall"),
                                used_var),
                   data = micro2_42h_df)

# Create an lm model for emmeans compatibility
lm_model <- lm(
  reformulate(c("DEPTH", "Rainfall", "DEPTH:Rainfall"), used_var),
  data = micro2_42h_df
)
anova_summary <- summary(anova_model)

# Compute emmeans and pairwise comparisons
emm_depth <- emmeans(lm_model, ~ DEPTH)
emm_rain <- emmeans(lm_model, ~ Rainfall)

# Pairwise comparisons with Tukey adjustment
pairs_depth <- pairs(emm_depth, adjust = "tukey")
pairs_rain <- pairs(emm_rain, adjust = "tukey")

# save everything as a text file
sink(file.path(stats_dir, "anova_summary_micro2_42h.txt"))

cat("=== ANOVA Summary ===\n")
print(anova_summary)

cat("\n=== Post Hoc: DEPTH ===\n")
print(emm_depth)
cat("\n--- Pairwise Comparisons (DEPTH, Tukey) ---\n")
print(pairs_depth)

cat("\n=== Post Hoc: Rainfall ===\n")
print(emm_rain)
cat("\n--- Pairwise Comparisons (Rainfall, Tukey) ---\n")
print(pairs_rain)

sink()
```
