---
title: "micro_Nelson_calcs"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

This code works up litter moisture data from the 2022 IRES Microclimate project, both microphase 1 and 2. Original csv file is pulled in, cleaned, analyzed, and graphed.

File written: 13 July 2025, Heather Throop\
Last updated: 2026-01-30 HT - updated file location and directory structure

## Preparation Steps

```{r}
#| echo: false
#| label: load packages 

library(tidyverse)

sessionInfo()
```

```{r}
# Use the stored config file to define the path for the DERT Data & Metadata folder (in ASU Dropbox) 
DERT_Archives <- config::get()

# Base project paths
data_dir   <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "data")

icon_dir   <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "icons")

stats_dir  <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "output", "stats")

fig_dir    <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "output", "figures")
```

```{r}
#| label: read in data
  
micro1_weather_df <- read_csv(file.path(data_dir,"microphase1_L2_weather_data.csv"))
micro2_weather_df <- read_csv(file.path(data_dir,"microphase2_L2_weather_data.csv"))
micro1_litter_df <- read_csv(file.path(data_dir,"microphase1_L2_data.csv"))
micro2_litter_df <- read_csv(file.path(data_dir,"microphase2_L2_data.csv"))

# change litter moisture on a percent basis (LitterMoisturePct) to g/g basis (LitterMoisture) for consistency with Nelson's equation
micro1_litter_df <- micro1_litter_df |>
  mutate(LitterMoisture = LitterMoisturePct / 100)
micro2_litter_df <- micro2_litter_df |>
  mutate(LitterMoisture = LitterMoisturePct / 100)

# change AirTemp to AirTemperature for consistency with micro1 dataframes
micro2_weather_df <- micro2_weather_df |>
  rename(AirTemperature = AirTemp)

```

delG = -(RT/M)*ln(H/100)
R = 1.987 (constant) # universal gas constant in cal/(mol*K)
T = temperature in Kelvin 
M = 18.01528 (constant) # molar mass of water in g/mol
H = relative humidity (%) 


```{r}
#| label: Nelson-eq-2

# calculate Gibbs free energy (delG) for water vapor condensation using Equation 2 from Nelson

R <- 1.987 # universal gas constant in cal/(mol*K)
M <- 18.01528 # molar mass of water in g/mol

micro1_weather_df <- micro1_weather_df |>
  mutate(
    T_K = AirTemperature + 273.15, # convert temperature to Kelvin
    delG = -(R * T_K / M) * log(Humidity / 100), # calculate delG in cal/mol
    ln_delG = log(delG) # calculate the natural log of delG 
  )

micro2_weather_df <- micro2_weather_df |>
  mutate(
    T_K = AirTemperature + 273.15, # convert temperature to Kelvin
    delG = -(R * T_K / M) * log(Humidity / 100), # calculate delG in cal/mol
    ln_delG = log(delG) # calculate the natural log of delG 
  )

```

## Micro1 Data

```{r}

# Left join micro1_weather_df and micro1_means
# Filter to keep only rows where Rainfall equals 0
micro1_nelson_df <- micro1_weather_df |>
  left_join(
    micro1_litter_df |>
      filter(Rainfall == 0),
    by = "elapsed_time"
  ) |>
  filter(!is.na(Species))

```

```{r}
# Create micro1 scatter plot
ggplot(micro1_nelson_df, aes(x = Humidity, y = LitterMoisture)) +
  geom_point() +
  facet_grid(Species ~ Littercond) +
  labs(
    title = "Microphase 1",
    x = "Relative Humidity (%)",
    y = "Litter Moisture (g g-1)",
    caption = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "lightgrey"),
    strip.text = element_text(face = "bold")
  )
```

## Micro2 Data


```{r}

# Left join micro2_weather_df and micro2_means
# Filter to keep only rows where Rainfall equals 0, litter is on the surface, and to remove extra met station rows (e.g., Gobabeb data when the station was not recording)
micro2_nelson_df <- micro2_weather_df |>
  left_join(
    micro2_litter_df |>
      filter(Rainfall == 0)) |>
      filter(DEPTH == 0) |>
      filter(!is.na(Rainfall)) |>
      filter(!is.na(AirTemperature))

```

```{r}
# Create micro2 scatter plot
ggplot(micro2_nelson_df, aes(x = Humidity, y = LitterMoisture)) +
  geom_point() +
#  facet_grid(Species ~ Littercond) +
  labs(
    title = "Microphase 2",
    x = "Relative Humidity (%)",
    y = "Litter Moisture (g g-1)",
    caption = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "lightgrey"),
    strip.text = element_text(face = "bold")
  )
```

```{r}
#| label: nelson-regression
  
  # Perform linear regression to find A and B
  model <- lm(ln_delG ~ LitterMoisture, data = micro2_nelson_df)
  
  # Extract A (intercept) and B (slope)
  A <- coef(model)[1]
  B <- coef(model)[2]
  
  # Print results
  cat("Linear Regression Results:\n")
  cat("A (intercept) =", A, "\n")
  cat("B (slope) =", B, "\n")
  cat("ln(G0) =", A, "\n")
  cat("G0 =", exp(A), "\n\n")
  
  # Print model summary for additional statistics
  print(summary(model))
  
  # Create a plot to visualize the fit
  plot(micro2_nelson_df$LitterMoisture, micro2_nelson_df$ln_delG, 
       main = "Linear Regression: ln(G) vs m",
       xlab = "m", ylab = "ln(G)",
       pch = 16)
  abline(model, col = "red")
  
```


```{r}
# Function to calculate and plot the relationship between m and H
plot_m_vs_H <- function(A, B, T_value) {
  # Constants
  R <- 1.987
  M <- 18
  
  # Function to calculate m from H and T using the derived equation
  calculate_m <- function(H, T, A, B) {
    # Calculate the inner term
    inner_term <- -(R * T / (M * exp(A))) * log(H/100)
    
    # Calculate m
    m <- (1/B) * log(inner_term)
    return(m)
  }
  
  # Create a sequence of H values for plotting (from 1% to 99%)
  H_values <- seq(1, 99, by = 0.5)
  
  # Calculate corresponding m values
  m_values <- sapply(H_values, function(h) {
    tryCatch({
      calculate_m(h, T_value, A, B)
    }, error = function(e) {
      NA  # Return NA for values that cause errors (e.g., log of negative numbers)
    })
  })
  
  # Remove any NA values that might have occurred due to mathematical constraints
  valid_indices <- !is.na(m_values)
  H_valid <- H_values[valid_indices]
  m_valid <- m_values[valid_indices]
  
  # Create the plot
  plot(H_valid, m_valid, 
       type = "l", 
       col = "blue", 
       lwd = 2,
       main = paste("Relationship Between H and m at T =", T_value, "K"),
       xlab = "Relative Humidity H (%)",
       ylab = "m",
       xlim = c(min(H_valid), max(H_valid)),
       ylim = c(min(m_valid, na.rm = TRUE), max(m_valid, na.rm = TRUE)))
  
  # Add grid for better readability
  grid()
  
  # Add a note about the equation used
  mtext(paste("Using: m = (1/B) * ln(-(R*T/(M*exp(A))) * ln(H/100))"), 
        side = 3, 
        line = 0.5, 
        cex = 0.8)
  
  # Return the data used for plotting
  return(data.frame(H = H_valid, m = m_valid))
}

# Complete workflow function that performs regression and then plots
analyze_and_plot <- function(data_df, T_for_plot) {
  # Calculate G using the second equation
  calculate_G <- function(T, H) {
    R <- 1.987
    M <- 18
    G <- -(R * T / M) * log(H/100)
    return(G)
  }
  
  # Calculate G and ln(G)
  data_df$G <- calculate_G(data_df$T, data_df$H)
  data_df$ln_G <- log(data_df$G)
  
  # Perform linear regression to find A and B
  model <- lm(ln_G ~ m, data = data_df)
  
  # Extract A (intercept) and B (slope)
  A <- coef(model)[1]
  B <- coef(model)[2]
  
  # Print results
  cat("Linear Regression Results:\n")
  cat("A (intercept) =", A, "\n")
  cat("B (slope) =", B, "\n")
  cat("ln(G0) =", A, "\n")
  cat("G0 =", exp(A), "\n\n")
  
  # Print model summary
  print(summary(model))
  
  # Create regression plot
  par(mfrow = c(1, 2))  # Set up for two plots side by side
  
  # Plot 1: Linear regression
  plot(data_df$m, data_df$ln_G, 
       main = "Linear Regression: ln(G) vs m",
       xlab = "m", ylab = "ln(G)",
       pch = 16)
  abline(model, col = "red")
  
  # Plot 2: m vs H relationship
  plot_data <- plot_m_vs_H(A, B, T_for_plot)
  
  # Reset plot layout
  par(mfrow = c(1, 1))
  
  # Return all results
  return(list(
    model = model, 
    A = A, 
    B = B, 
    G0 = exp(A),
    plot_data = plot_data
  ))
}
```

