---
title: "IRES_microclimate_Nelson_calcs"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

This code explores the Nelson 1984 equation predictions for equilibrium litter moisture content relative to data collect in the field from the 2022 IRES Microclimate project. Data from both microphase 1 and 2 experiments are used. 

File written: 2026-04-04, Heather Throop\
Last updated: 2026-04-19 HT, added code to calculate A and B values for each species and litter condition combination  and to plot the resulting Gibbs equation curves on scatter plots. 

## Preparation Steps

```{r}
#| echo: false
#| label: load packages 

library(tidyverse)

sessionInfo()
```

```{r}
# Use the stored config file to define the path for the DERT Data & Metadata folder (in HT's ASU Dropbox) 
DERT_Archives <- config::get()

# Base project paths
data_dir   <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "data")

stats_dir  <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "output", "stats")

fig_dir    <- file.path(DERT_Archives,
                        "NSF_IRES", "2022_Littermicroclimate",
                        "HT_compiled", "output", "figures")
```

```{r}
#| label: read in data

# weather_df files are meterological data collected at the met station, litter_df files are litter moisture data collected from litter moisture manipulation experiments 
micro1_weather_df <- read_csv(file.path(data_dir,"microphase1_L2_weather_data.csv"))
micro2_weather_df <- read_csv(file.path(data_dir,"microphase2_L2_weather_data.csv"))
micro1_litter_df <- read_csv(file.path(data_dir,"microphase1_L2_data.csv"))
micro2_litter_df <- read_csv(file.path(data_dir,"microphase2_L2_data.csv"))

# change litter moisture on a percent basis (LitterMoisturePct) to g/g basis (LitterMoisture) for consistency with Nelson's equation
micro1_litter_df <- micro1_litter_df |>
  mutate(LitterMoisture = LitterMoisturePct / 100)
micro2_litter_df <- micro2_litter_df |>
  mutate(LitterMoisture = LitterMoisturePct / 100)

# change AirTemp to AirTemperature for consistency with micro1 dataframes
micro2_weather_df <- micro2_weather_df |>
  rename(AirTemperature = AirTemp)

# read in the isotherm parameters from Anderson 1990 
Anderson_values <- read_csv(file.path(data_dir, "Anderson1990_data.csv"))

```

Nelson 1984 equation 2 for Gibbs free energy of water vapor condensation on litter surfaces:
delG = -(RT/M)*ln(H/100)
R = 1.987 (constant) # universal gas constant in cal/(mol*K)
T = temperature in Kelvin 
M = 18.01528 (constant) # molar mass of water in g/mol
H = relative humidity (%) 

```{r}
#| label: Nelson-eq-2

# define constants for Nelson equation 2

R <- 1.987 # universal gas constant in cal/(mol*K)
M <- 18.01528 # molar mass of water in g/mol

```

## Microphase1 Experiment

Data are from the first IRES Microclimate field experiment in 2022. Litter packets from two species (Stipagrostis sabulicula and Salvadora persica) and two litter conditions (green and brown) were placed on the soil surface and subjected to one of three watering treatments (0, 5, 10 mm). Gravimetic moisture was measured periodically by destructively harvesting samples. Only the surface litter packets that received no added rainfall are used in this analysis.

### calculations

```{r}
#| label: Microphase1 data prep for Nelson equation

# calculate Gibbs free energy (delG) for water vapor condensation using Equation 2 from Nelson for microphase 1 data
micro1_weather_df <- micro1_weather_df |>
  mutate(
    T_K = AirTemperature + 273.15, # convert temperature to Kelvin
    delG = -(R * T_K / M) * log(Humidity / 100), # calculate delG in cal/mol
    ln_delG = log(delG) # calculate the natural log of delG 
  )

# Left join micro1_weather_df and micro1_means
# Filter to keep only rows where Rainfall equals 0 - these are the litter packets that did not receive added rainfall (but still experienced fog)
micro1_nelson_df <- micro1_weather_df |>
  left_join(
    micro1_litter_df |>
      filter(Rainfall == 0),
    by = "elapsed_time"
  ) |>
  filter(!is.na(Species))

```

```{r}
#| label: Microphase1 scatter plot 

# Create micro1 scatter plot - litter moisture vs humidity, faceted by species and litter condition
p_micro1 <- ggplot(micro1_nelson_df, aes(x = Humidity, y = LitterMoisture)) +
  geom_point() +
  facet_grid(Species ~ Littercond) +
  labs(
  #  title = "Microphase 1",
    x = "Relative Humidity (%)",
    y = expression("Litter Moisture (g g"^{-1}*")"),
    caption = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    strip.background = element_rect(fill = "lightgrey"),
    strip.text = element_text(face = "bold")
  )
p_micro1
```

```{r}
#| label: calculate A and B values for each species and litter condition combination

# These are the isotherm parameters from Nelson 1984, A is the intercept and B is the slope of the 
# linear regression of ln(G) vs m (litter moisture content in g/g)
# Nelson assumes that temperature is constant, which in reality it was not. 

# Create a dataframe to store A and B values for each combination
ab_values <- data.frame()

# Get unique combinations of Species and Littercond
unique_combos <- unique(micro1_nelson_df[, c("Species", "Littercond")])

# For each unique Species and Littercond combination
for(i in 1:nrow(unique_combos)) {
  current_species <- unique_combos$Species[i]
  current_littercond <- unique_combos$Littercond[i]
  
  # Filter data for this combination
  subset_data <- micro1_nelson_df[
    micro1_nelson_df$Species == current_species & 
    micro1_nelson_df$Littercond == current_littercond, ]
  
  # Skip if there's not enough data
  if(nrow(subset_data) < 3) {
    cat("Skipping", current_species, "-", current_littercond, "due to insufficient data\n")
    next
  }
  
  # Perform linear regression to find A and B
  model <- lm(ln_delG ~ LitterMoisture, data = subset_data)
  
  # Extract A (intercept) and B (slope)
  A_val <- coef(model)[1]
  B_val <- coef(model)[2]
  
  # Store results
  ab_values <- rbind(ab_values, data.frame(
    Species = current_species,
    Littercond = current_littercond,
    A = A_val,
    B = B_val,
    G0 = exp(A_val),
    R_squared = summary(model)$r.squared
  ))
  
  # Print results for this combination
  cat("\n=== Results for", current_species, "-", current_littercond, "===\n")
  cat("A (intercept) =", A_val, "\n")
  cat("B (slope) =", B_val, "\n")
  cat("ln(G0) =", A_val, "\n")
  cat("G0 =", exp(A_val), "\n")
  cat("R-squared =", summary(model)$r.squared, "\n")
  
  # Create a plot for this combination
  plot_title <- paste("Linear Regression:", current_species, "-", current_littercond)
  plot(subset_data$LitterMoisture, subset_data$ln_delG, 
       main = plot_title,
       xlab = "Litter Moisture (g g-1)", ylab = "ln(G)",
       pch = 16)
  abline(model, col = "red")
}

# Print the final table of A and B values
print(ab_values)
```

The intercepts are reasonable relative to other values but the slopes is less negative than typical from the literature. This seems to be due to the high range in litter moisture values. Should these be filtered to only include the lowest litter moisture values when ln(G) is -1??

### plot Nelson curves

```{r}
#| label: plot Nelson curves (using litter-specific parameters) versus m1 data

# parameters A and B obtained from the linear regression of ln(G) vs m for microphase 1 data - likely not at equilibrium

# Create an empty list to store dataframes for each combination
gibbs_curves_list <- list()

# For each Species and Littercond combination with calculated A and B
for(i in 1:nrow(ab_values)) {
  # Extract the specific A and B values
  current_species <- ab_values$Species[i]
  current_littercond <- ab_values$Littercond[i]
  current_A <- ab_values$A[i]
  current_B <- ab_values$B[i]
  
# Get average air temperature for the duration of the experiment
meanT_micro1 <- mean(micro1_nelson_df$AirTemperature, na.rm = TRUE)
  
  # Create a dataframe with humidity range present during this experiment (25-100%)
  temp_df <- data.frame(
    Humidity = seq(25, 100, by = 0.5),
    Species = current_species,
    Littercond = current_littercond
  )
  
  # Calculate Gibbs equation values using the specific A and B values for each species x littercond combination
  temp_df$GibbsEqn_m <- (1/current_B) * 
    (log(((-R*(meanT_micro1 + 273.15))/(M*exp(current_A))) * log(temp_df$Humidity/100)))
  
  # Add to the list
  gibbs_curves_list[[i]] <- temp_df
}

# Combine all dataframes
gibbs_curves_df <- do.call(rbind, gibbs_curves_list)

# Add the Gibbs equation lines to the plot
p_micro1 + geom_line(data = gibbs_curves_df, 
              aes(x = Humidity, y = GibbsEqn_m), 
              color = "blue") +
  labs(title = "Microphase 1 - species and litter condition isotherm parameters used")
```

```{r}
#| label: plot Nelson curves (literature data) versus cheatgrass data 

# Gibbs A and B values obtained from lab studies (Anderson, 1990) 
# values extracted from Kathe's code
A_cheatgrass <- 4.9145
B_cheatgrass <- -18.15

# Create an empty list to store dataframes for each combination
gibbs_curves_list <- list()

# Get unique combinations of Species and Littercond
unique_combos <- unique(micro1_nelson_df[, c("Species", "Littercond")])

# For each Species and Littercond combination
for(i in 1:nrow(unique_combos)) {
  # Extract the current species and littercond
  current_species <- unique_combos$Species[i]
  current_littercond <- unique_combos$Littercond[i]
 
  # Get average air temperature for all data
  meanT_micro1 <- mean(micro1_nelson_df$AirTemperature, na.rm = TRUE)
  
  # Create a dataframe with humidity range 25-100
  temp_df <- data.frame(
    Humidity = seq(25, 100, by = 0.5),
    Species = current_species,
    Littercond = current_littercond
  )
  
  # Calculate Gibbs equation values using the stored A and B
  temp_df$GibbsEqn_m <- (1/B_cheatgrass) * 
    (log(((-R*(meanT_micro1 + 273.15))/(M*exp(A_cheatgrass))) * log(temp_df$Humidity/100)))
  
  # Add to the list
  gibbs_curves_list[[i]] <- temp_df
}

# Combine all dataframes
gibbs_curves_cheatgrass_m1_df <- do.call(rbind, gibbs_curves_list)

# Add the Gibbs equation lines to the plot
p_micro1 + geom_line(data = gibbs_curves_cheatgrass_m1_df, 
              aes(x = Humidity, y = GibbsEqn_m), 
              color = "blue")+
  labs(title = "Microphase 1 - Cheatgrass isotherm parameters")
```

```{r}
#| label: plot Nelson curves versus Anderson data for micro1 - cheatgrass and quaking aspen

# Filter Anderson_values to only use Temperature = 278 and AD_DE = "AD"
anderson_filtered <- Anderson_values[Anderson_values$Temperature == 278 & 
                                    Anderson_values$AD_DE == "AD", ]

# Create an empty list to store dataframes for each combination
gibbs_curves_list <- list()

# Get unique combinations of Species and Littercond
unique_combos <- unique(micro1_nelson_df[, c("Species", "Littercond")])

# For each Species and Littercond combination
for(i in 1:nrow(unique_combos)) {
  # Extract the current species and littercond
  current_species <- unique_combos$Species[i]
  current_littercond <- unique_combos$Littercond[i]
  
  # Map the species and littercond to Anderson values
  anderson_species <- ifelse(current_species == "grass", "Cheatgrass", 
                           ifelse(current_species == "shrub", "QuakingAspen", NA))
  
  anderson_littercond <- ifelse(current_littercond == "Brown", "weathered",
                              ifelse(current_littercond == "Green", "fresh", NA))
  
  # Skip if we don't have a valid mapping
  if(is.na(anderson_species) || is.na(anderson_littercond)) {
    next
  }
  
  # Find the matching row in anderson_filtered
  # Make sure we're using column names that exist in the dataset
  matching_row <- anderson_filtered[anderson_filtered$Species == anderson_species & 
                                   anderson_filtered$Littercond == anderson_littercond, ]
  
  # If we found a match, use those A and B values
  if(nrow(matching_row) > 0) {
    current_A <- matching_row$A[1]  # Take the first match if multiple
    current_B <- matching_row$B[1]
    
    # Get average air temperature for all data
    meanT_micro1 <- mean(micro1_nelson_df$AirTemperature, na.rm = TRUE)
    
    # Create a dataframe with humidity range 25-100
    temp_df <- data.frame(
      Humidity = seq(25, 100, by = 0.5),
      Species = current_species,
      Littercond = current_littercond
    )
    
    # Calculate Gibbs equation values using the matched A and B
    temp_df$GibbsEqn_m <- (1/current_B) * 
      (log(((-R*(meanT_micro1 + 273.15))/(M*exp(current_A))) * log(temp_df$Humidity/100)))
    
    # Add to the list
    gibbs_curves_list[[i]] <- temp_df
  } # Close the if(nrow(matching_row) > 0) block
} # Close the for loop

# Combine all dataframes (only if there are any valid matches)
if(length(gibbs_curves_list) > 0) {
  gibbs_curves_df <- do.call(rbind, gibbs_curves_list)
  
  # Create and return the plot
  m1_plot_Nelson <- p_micro1 + 
    geom_line(data = gibbs_curves_df, 
              aes(x = Humidity, y = GibbsEqn_m), color = "blue") #+
   # labs(title = "Microphase 1 - Isotherm parameters (Anderson 1990, T=278K)") 

  # Return the plot
  m1_plot_Nelson
} else {
  # If no matches found, return the original plot with a warning message
  warning("No matching A and B values found in Anderson_values dataset")
  p_micro1
}

# save the plot
ggsave(filename = file.path(DERT_Archives, "NSF_IRES", "2022_Littermicroclimate", 
                            "HT_compiled", "output", "figures", "Nelson_micro1.jpg"),
      plot = m1_plot_Nelson,
       width = 8, height = 6, dpi = 300)
```



## Microphase2 Experiment

Data are from the second IRES Microclimate experiment in 2022. Litter packets were placed on the soil surface and buried at two depths (1 and 3 cm) and subjected to watering treatments (0, 5, 10 mm). There was only one litter species and condition (brown grass). Gravimetic moisture was measured periodically by destructively harvesting samples. Only the surface litter packets that received no added rainfall are used in this analysis.

### calculations

```{r}
#| label: m2 data prep for Nelson equation

# calculate Gibbs free energy (delG) for water vapor condensation using Equation 2 from Nelson for microphase 1 data
micro2_weather_df <- micro2_weather_df |>
  mutate(
    T_K = AirTemperature + 273.15, # convert temperature to Kelvin
    delG = -(R * T_K / M) * log(Humidity / 100), # calculate delG in cal/mol
    ln_delG = log(delG) # calculate the natural log of delG 
  )

# Left join micro2_weather_df and micro2_means
# Filter to keep only rows where Rainfall equals 0, litter is on the surface, and to remove extra met station rows (e.g., Gobabeb data when the station was not recording)
micro2_nelson_df <- micro2_weather_df |>
  left_join(
    micro2_litter_df |>
      filter(Rainfall == 0)) |>
      filter(DEPTH == 0) |>
      filter(!is.na(Rainfall)) |>
      filter(!is.na(AirTemperature))

```

```{r}
#| label: Create micro2 scatter plot

p_micro2 <- ggplot(micro2_nelson_df, aes(x = Humidity, y = LitterMoisture)) +
  geom_point() +
  labs(
  #  title = "Microphase 2",
    x = "Relative Humidity (%)",
    y = expression("Litter Moisture (g g"^{-1}*")"),
    caption = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    strip.background = element_rect(fill = "lightgrey"),
    strip.text = element_text(face = "bold")
  )

p_micro2
```

```{r}
#| label: nelson-regression
  
  # Perform linear regression to find A and B
  model <- lm(ln_delG ~ LitterMoisture, data = micro2_nelson_df)
  
  # Extract A (intercept) and B (slope)
  A <- coef(model)[1]
  B <- coef(model)[2]
  
  # Print results
  cat("Linear Regression Results:\n")
  cat("A (intercept) =", A, "\n")
  cat("B (slope) =", B, "\n")
  cat("ln(G0) =", A, "\n")
  cat("G0 =", exp(A), "\n\n")
  
  # Print model summary for additional statistics
  print(summary(model))
  
  # Create a plot to visualize the fit
  plot(micro2_nelson_df$LitterMoisture, micro2_nelson_df$ln_delG, 
       main = "Linear Regression: ln(G) vs m",
       xlab = "m", ylab = "ln(G)",
       pch = 16)
  abline(model, col = "red")
  
```

Note that for this experiment the highest litter moisture value is <0.2 g/g, which is much lower than the range of litter moisture values in the microphase 1 experiment. 

### plot Nelson curves

```{r}
#| label: plot Nelson curves (using litter-specific parameters) versus m2 data

# parameters A and B obtained from the linear regression of ln(G) vs m for microphase 2 data - likely not at equilibrium

# calculate Gibbs free energy (delG) for water vapor condensation using Equation 2 from Nelson for microphase 2 data
micro2_nelson_df <- micro2_nelson_df |>
  mutate(
#Gibbs equilibrium moisture content
    GibbsEqn_m = (1/B)*
        (log(((-R*(AirTemperature + 273.15))/(M*exp(A)))* log(Humidity/100)))
  )

# calculate mean air temperature for the study period
meanT_micro2 <- mean(micro2_nelson_df$AirTemperature, na.rm = TRUE)

# Create a new dataframe with x values from 50 to 100
gibbs_curve_df <- data.frame(Humidity = seq(50, 100, by = 0.5))

# Calculate Gibbs equation values for the range of RH values present in the data, using mean air temperature for the study period (the equation is not very sensitive to temperature within the range of study conditions)
gibbs_curve_df <- gibbs_curve_df |>
  mutate(
    GibbsEqn_m = (1/B)*
      (log(((-R*(meanT_micro2 + 273.15))/(M*exp(A)))* log(Humidity/100)))
  )

# Add the Nelson equation line to the Microphase2 plot
p_micro2 + 
  geom_line(data = gibbs_curve_df, aes(x = Humidity, y = GibbsEqn_m), color = "blue", size = 1) +
  labs(title = "Microphase 2 - Cheatgrass isotherm parameters")

```

```{r}
#| label: plot Nelson curves (literature data) versus m2 data 

# Calculate Nelson equation values for the range of RH values present in the data, using mean air temperature for the study period (the equation is not very sensitive to temperature within the range of study conditions)

#assign values for A and B for cheatgrass from Anderson 1990. This is the weathered adsorbing material at 278K.

A_cheatgrass <- 4.945
B_cheatgrass <- -16.72

gibbs_curve_m2_cheatgrass_df <- gibbs_curve_df |>
  mutate(
    GibbsEqn_m = (1/B_cheatgrass)*
      (log(((-R*(meanT_micro2 + 273.15))/(M*exp(A_cheatgrass)))* log(Humidity/100)))
  )

# Add the Gibbs equation lines to the plot
Nelson_m2 <- p_micro2 + geom_line(data = gibbs_curve_m2_cheatgrass_df, 
              aes(x = Humidity, y = GibbsEqn_m), 
              color = "blue") #+
 # labs(title = "Microphase 2 - Cheatgrass isotherm parameters")
Nelson_m2

# save the plot
ggsave(filename = file.path(DERT_Archives, "NSF_IRES", "2022_Littermicroclimate", 
                            "HT_compiled", "output", "figures", "Nelson_micro2.jpg"),
      plot = Nelson_m2,
       width = 8, height = 6, dpi = 300)
```




